### 2025-05-28 chapters/haiti/haiti.Rnw
Fixed double author-year citations (manually)

The discrepancy between the model-based conclusions of \citet{lee20} and the prolonged absence of cholera in Haiti has been debated \citep{francois20,rebaudetComment20,henrys20,leeReply20}.
Suggested origins of this discrepancy include the use of unrealistic models \citep{rebaudetComment20} and unrealistic criteria for cholera elimination \citep{henrys20}.
We find a more nuanced conclusion: attention to methodological details in model fitting, diagnosis and forecasting can improve each of the proposed model's ability to quantitatively describe observed data.
This improved ability may result in more accurate forecasts and facilitates the exploration of model assumptions.
Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.

@@

In this chapter, we refer to models that focus on learning relationships between variables in a dataset as {\it associative}, whereas models that incorporate a known scientific property of the system we call {\it causal} or {\it mechanistic}.
The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
\citet{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
To successfully predict the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
This motivates the development of mechanistic models, which provides a statistical fit to the available data while also supporting a causal interpretation.
Despite their limited ability to project the effect of interventions on a system, associative models can be effectively used to make inference on an certain features of a system.
In our literature review, there were 14 studies that used associative models to describe various aspects of the cholera epidemic \citep{hulland19,moise20,piarroux11,matias17,eisenberg13,rebaudet19CATI,ivers15,raila17,cuneo17,charles14,richterman19,bengtsson15,richterman19inverse,li16adjusting}.

>>>>

The discrepancy between the model-based conclusions of \citet{lee20} and the prolonged absence of cholera in Haiti has been debated \citep[Lee et al., 2020a;][]{francois20,rebaudetComment20,henrys20}. \nocite{leeReply20}
Suggested origins of this discrepancy include the use of unrealistic models \citep{rebaudetComment20} and unrealistic criteria for cholera elimination \citep{henrys20}.
We find a more nuanced conclusion: attention to methodological details in model fitting, diagnosis and forecasting can improve each of the proposed model's ability to quantitatively describe observed data.
This improved ability may result in more accurate forecasts and facilitates the exploration of model assumptions.
Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.

@@

In this chapter, we refer to models that focus on learning relationships between variables in a dataset as {\it associative}, whereas models that incorporate a known scientific property of the system we call {\it causal} or {\it mechanistic}.
The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
\citet{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
To successfully predict the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
This motivates the development of mechanistic models, which provides a statistical fit to the available data while also supporting a causal interpretation.
Despite their limited ability to project the effect of interventions on a system, associative models can be effectively used to make inference on an certain features of a system.
In our literature review, there were 14 studies that used associative models to describe various aspects of the cholera epidemic \citep[Richterman et al., 2019a, 2019b;][]{hulland19,moise20,piarroux11,matias17,eisenberg13,rebaudet19CATI,ivers15,raila17,cuneo17,charles14,bengtsson15,li16adjusting}. \nocite{richterman19a}\nocite{richterman19b}
